Drought vulnerability of tree growth

### 2024-07-09 scripts/explore_analysis.Rmd
updated script

#first fitting a line through the time series
obs.model.dbh<-nlme::lme(calcDBH~Cno, random=~1|Tag, data=dendro.data,
                         na.action=na.omit,
                         method="ML", control=(msMaxIter=100)) #had to increase the number of iterations

#standardised residuals calculated https://stat.ethz.ch/R-manual/R-devel/library/nlme/html/residuals.lme.html

std.res.dbh<-residuals(obs.model.dbh, type="pearson")
hist(std.res.dbh)
dendro.data$std.res<-rep(NA, nrow(dendro.data))
dendro.data$std.res[!is.na(dendro.data$calcDBH)]<-std.res.dbh
#defining all observation with abs(residual) >3 as an outlier
dendro.data$out<-factor(ifelse(abs(dendro.data$std.res)<=3, 0, 1))

#find the trees which have outlier observations
outs.Tag<-unique(dendro.data$Tag[dendro.data$out==1])

#plot trees that have outliers
ggplot(dendro.data[dendro.data$Tag %in%outs.Tag,],
       aes(x = Cno, y = calcDBH, col=factor(out), group = Tag))+
  geom_line(alpha=0.5)+geom_point()+
  scale_color_manual(values=c("grey20", "red", "grey"))

dendro.data.noouts<-dendro.data %>%
  filter(abs(std.res)<=3)%>%
  select(-c(std.res, out))
  #mutate(dbh3=ifelse(abs(std.res)<=3, calcDBH, NA)) #%>% #create an outlier removed column

@@

#add long-term increment to dendro data
dendro.data.noouts<-merge(dendro.data.noouts, hkk_census, by="Tag", all.x=T)
dendro.data<-merge(dendro.data, hkk_census, by="Tag", all.x=T)

@@

dendro_inc<- dendro.data %>%
  filter(Cno %in% alt_census)%>%
  arrange(Cno)%>%
  group_by(Tag) %>% #grouping by treeID to calculate increment
  dplyr::mutate(calcDBH_min1=dplyr::lag(calcDBH, n=1, default=NA, order_by=Tag),
         cii_min1=dplyr::lag(Cii, n=1, default=NA, order_by=Tag),
         tdif=difftime(Date2, dplyr::lag(Date2, n=1, default=NA, order_by=Tag), units = "days"),
         inc=calcDBH-calcDBH_min1,
         inc_annual=(inc/as.numeric(tdif))*365,
         inc.time=Cno-dplyr::lag(Cno, n=1, default=NA, order_by=Tag),
         #removing increment value if the previous measure was before the last year
         inc_annual=ifelse(inc.time==2, inc_annual, NA),
         treeID=paste0("HKK.", Tag))%>%
  filter(!is.na(inc))%>%
  filter(!is.na(inc_annual))%>%
  ungroup()%>%
  dplyr::mutate(
    diainc.scaled = scale(inc_annual),
    out.inc = ifelse(abs(diainc.scaled)>3, 1, 0))%>%
  filter(out.inc==0)
```
```{r species wise increments, echo=F}

@@

@@

spagplot_top4<-ggplot() +
  #species plots
  geom_line(data=top4 %>% group_by(Cno, sp) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), 
  aes(x=((Cno-3)/2)+2009, y=mean_inc, 
  group=factor(sp,levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")), 
  col=factor(sp,levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))))+
  geom_line(data=top4 %>% group_by(Cno) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), 
  aes(x=((Cno-3)/2)+2009, y=mean_inc), col="black", size=2)+
  scale_color_viridis_d()+
  guides(col=guide_legend(title="species"))+
  xlab("year") + ylab("diameter increment (cm)")+
  ggtitle("growth increments for top 4 species")+
  theme_bw()

@@

@@

spagplot_top10<-ggplot()+
  #species plots
  geom_line(data=dendro_inc %>% filter(sp %in% sp_top10)%>% group_by(Cno, sp) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), 
  aes(x=((Cno-3)/2)+2009, y=mean_inc, 
  group=sp, col=sp))+
  #mean of all trees
  geom_line(data=dendro_inc %>% filter(sp %in% sp_top10)%>% group_by(Cno) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), 
  aes(x=((Cno-3)/2)+2009, y=mean_inc), col="black", size=2)+
  #mean of species
  geom_line(data=dendro_inc %>% filter(sp %in% sp_top10)%>% group_by(Cno, sp) %>% 
  dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)) %>%
  ungroup()%>%
  group_by(Cno) %>% dplyr::summarise(mean_inc=mean(mean_inc, na.rm=T)), 
  aes(x=((Cno-3)/2)+2009, y=mean_inc), col="grey40", size=0.8)+
  scale_color_viridis_d()+
  guides(col=guide_legend("species"))+
  xlab("year") + ylab("diameter increment (cm)")+
  ggtitle("growth increments for top 10 species")+
  theme_bw()

@@

spagplot_all<-ggplot() +
  #species plots
  geom_line(data=dendro_inc %>% group_by(Cno, sp) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), 
  aes(x=((Cno-3)/2)+2009, y=mean_inc, 
  group=sp, col=sp))+
  #mean of all trees
  geom_line(data=dendro_inc %>% group_by(Cno) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), 
  aes(x=((Cno-3)/2)+2009, y=mean_inc), col="black", size=2)+
  #mean of species
  geom_line(data=dendro_inc %>% group_by(Cno, sp) %>% 
  dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)) %>%
  ungroup()%>%
  group_by(Cno) %>% dplyr::summarise(mean_inc=mean(mean_inc, na.rm=T)), 
  aes(x=((Cno-3)/2)+2009, y=mean_inc), col="grey40", size=0.8)+
  scale_color_viridis_d()+
  guides(col="none", group="none")+
  xlab("year") + ylab("diameter increment (cm)")+
  ggtitle("growth increments for all species")+
  theme_bw()

@@

```
```{r top4 species growth contribution, echo=F}

>>>>

dbh.data<-dendrobyband %>%
  dplyr::rename(sp=SPCODE.UPDATE)%>%
  mutate(year=year(Date2)) %>%
  filter(is.na(dbh.flag), dbh.decimal.error=="none")%>%
  group_by(Cno, sp, Tag) %>%
  dplyr::summarise(dbh=DbhDendro[which(n.dendro==min(n.dendro))],
                   Date2=Date2[which(n.dendro==min(n.dendro))],
                   Cii=max(Cii, na.rm=T))

@@

#add long-term increment to dendro data
dendro.data<-merge(dendro.data, hkk_census, by="Tag", all.x=T)

@@

dendro_inc<- dendro.data %>%
  filter(Cno %in% alt_census)%>%
  arrange(Cno)%>%
  group_by(Tag) %>% #grouping by treeID to calculate increment
  dplyr::mutate(calcDBH_min1=dplyr::lag(calcDBH, n=1, default=NA, order_by=Tag),
         cii_min1=dplyr::lag(Cii, n=1, default=NA, order_by=Tag),
         tdif=difftime(Date2, dplyr::lag(Date2, n=1, default=NA, order_by=Tag), units = "days"),
         inc=calcDBH-calcDBH_min1,
         inc_annual=(inc/as.numeric(tdif))*365,
         inc.time=Cno-dplyr::lag(Cno, n=1, default=NA, order_by=Tag),
         #removing increment value if the previous measure was before the last year
         inc_annual=ifelse(inc.time==2, inc_annual, NA),
         treeID=paste0("HKK.", Tag))%>%
  filter(!is.na(inc))%>%
  filter(!is.na(inc_annual))%>%
  ungroup()%>%
  dplyr::mutate(
    diainc.scaled = scale(inc_annual),
    out.inc = ifelse(abs(diainc.scaled)>3, 1, 0))%>%
  filter(out.inc==0)

dbh_inc<- dbh.data %>%
  filter(Cno %in% alt_census)%>%
  arrange(Cno)%>%
  group_by(Tag) %>% #grouping by treeID to calculate increment
  dplyr::mutate(dbh_min1=dplyr::lag(dbh, n=1, default=NA, order_by=Tag),
         cii_min1=dplyr::lag(Cii, n=1, default=NA, order_by=Tag),
         tdif=difftime(Date2, dplyr::lag(Date2, n=1, default=NA, order_by=Tag), units = "days"),
         inc=dbh-dbh_min1,
         inc_annual=(inc/as.numeric(tdif))*365,
         inc.time=Cno-dplyr::lag(Cno, n=1, default=NA, order_by=Tag),
         #removing increment value if the previous measure was before the last year
         inc_annual=ifelse(inc.time==2, inc_annual, NA),
         treeID=paste0("HKK.", Tag))%>%
  filter(!is.na(inc))%>%
  filter(!is.na(inc_annual))%>%
  ungroup()%>%
  dplyr::mutate(
    diainc.scaled = scale(inc_annual),
    out.inc = ifelse(abs(diainc.scaled)>3, 1, 0))%>%
  filter(out.inc==0)

```
```{r species wise increments, echo=F}

@@

top4_dbh<-dbh_inc %>%
filter(sp %in% c("SACCLI", "HOPEOD", "TETRNU", "POLYVI"))

@@

spagplot_top4<-ggplot() +
  #species plots
  geom_line(data=top4 %>% group_by(Cno, sp) %>% dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), 
  aes(x=((Cno-3)/2)+2009, y=median_inc, 
  group=factor(sp,levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")), 
  col=factor(sp,levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))))+
  geom_line(data=top4 %>% group_by(Cno) %>% dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), 
  aes(x=((Cno-3)/2)+2009, y=median_inc), col="black", size=2)+
  scale_color_viridis_d()+
  guides(col=guide_legend(title="species"))+
  xlab("year") + ylab("diameter increment (cm)")+
  ggtitle("growth increments for top 4 species")+
  theme_bw()

spagplot_top4_dbh<-ggplot() +
  #species plots
  geom_line(data=top4_dbh %>% group_by(Cno, sp) %>% 
  dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), 
  aes(x=((Cno-3)/2)+2009, y=median_inc, 
  group=factor(sp,levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")), 
  col=factor(sp,levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))))+
  geom_line(data=top4_dbh %>% group_by(Cno) %>% dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), 
  aes(x=((Cno-3)/2)+2009, y=median_inc), col="black", size=2)+
  scale_color_viridis_d()+
  guides(col=guide_legend(title="species"))+
  xlab("year") + ylab("diameter increment (cm)")+
  ggtitle("growth increments for top 4 species")+
  theme_bw()

@@

#individual panels using facet_wrap
spagplot_dbh_ind<-ggplot() +
  geom_line(data = top4_dbh, aes(x=((Cno-3)/2)+2009, y=inc_annual, group=Tag), alpha=0.5) +
  geom_line(data=top4_dbh %>% group_by(Cno) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=mean_inc), col="red")+
  geom_line(data=top4_dbh %>% group_by(Cno) %>% dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=median_inc), col="blue")+
  facet_wrap(factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))~., 
  nrow=2)+
  xlab("year") + ylab("diameter increment (cm)")+
  ggtitle("growth increments for top 4 species from DBH measurements")+
  theme_bw()

png("growth-precip-thailand/doc/display/spaghetti_top4_dbh_ind.png", width=8, height=8, units="in", res=300)
spagplot_dbh_ind
dev.off()

png("growth-precip-thailand/doc/display/spaghetti_top4_dbh.png", width=4, height=4, units="in", res=300)
spagplot_top4_dbh
dev.off()

@@

spagplot_top10<-ggplot()+
  #species plots
  geom_line(data=dendro_inc %>% filter(sp %in% sp_top10)%>% group_by(Cno, sp) %>% 
  dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), 
  aes(x=((Cno-3)/2)+2009, y=median_inc, 
  group=sp, col=sp))+
  #mean of all trees
  geom_line(data=dendro_inc %>% filter(sp %in% sp_top10)%>% group_by(Cno) %>% 
  dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), 
  aes(x=((Cno-3)/2)+2009, y=median_inc), col="black", size=2)+
  #mean of species
  geom_line(data=dendro_inc %>% filter(sp %in% sp_top10)%>% group_by(Cno, sp) %>% 
  dplyr::summarise(median_inc=median(inc_annual, na.rm=T)) %>%
  ungroup()%>%
  group_by(Cno) %>% dplyr::summarise(median_inc=mean(median_inc, na.rm=T)), 
  aes(x=((Cno-3)/2)+2009, y=median_inc), col="grey40", size=0.8)+
  scale_color_viridis_d()+
  guides(col=guide_legend("species"))+
  xlab("year") + ylab("diameter increment (cm)")+
  ggtitle("growth increments for top 10 species")+
  theme_bw()

@@

spagplot_top10_dbh<-ggplot()+
  #species plots
  geom_line(data=dbh_inc %>% filter(sp %in% sp_top10)%>% group_by(Cno, sp) %>% 
  dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), 
  aes(x=((Cno-3)/2)+2009, y=median_inc, 
  group=sp, col=sp))+
  #mean of all trees
  geom_line(data=dbh_inc %>% filter(sp %in% sp_top10)%>% group_by(Cno) %>% 
  dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), 
  aes(x=((Cno-3)/2)+2009, y=median_inc), col="black", size=2)+
  #mean of species
  geom_line(data=dbh_inc %>% filter(sp %in% sp_top10)%>% group_by(Cno, sp) %>% 
  dplyr::summarise(median_inc=median(inc_annual, na.rm=T)) %>%
  ungroup()%>%
  group_by(Cno) %>% dplyr::summarise(median_inc=mean(median_inc, na.rm=T)), 
  aes(x=((Cno-3)/2)+2009, y=median_inc), col="grey40", size=0.8)+
  scale_color_viridis_d()+
  guides(col=guide_legend("species"))+
  xlab("year") + ylab("diameter increment (cm)")+
  ggtitle("growth increments for top 10 species")+
  theme_bw()

png("growth-precip-thailand/doc/display/spaghetti_top10_dbh.png", width=4, height=4, units="in", res=300)
spagplot_top10_dbh
dev.off()

spagplot_all<-ggplot() +
  #species plots
  geom_line(data=dendro_inc %>% group_by(Cno, sp) %>% 
  dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), 
  aes(x=((Cno-3)/2)+2009, y=median_inc, 
  group=sp, col=sp))+
  #mean of all trees
  geom_line(data=dendro_inc %>% group_by(Cno) %>% 
  dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), 
  aes(x=((Cno-3)/2)+2009, y=median_inc), col="black", size=2)+
  #mean of species
  geom_line(data=dendro_inc %>% group_by(Cno, sp) %>% 
  dplyr::summarise(median_inc=median(inc_annual, na.rm=T)) %>%
  ungroup()%>%
  group_by(Cno) %>% dplyr::summarise(median_inc=mean(median_inc, na.rm=T)), 
  aes(x=((Cno-3)/2)+2009, y=median_inc), col="grey40", size=0.8)+
  scale_color_viridis_d()+
  guides(col="none", group="none")+
  xlab("year") + ylab("diameter increment (cm)")+
  ggtitle("growth increments for all species")+
  theme_bw()

@@

#growth_top4_sum using dbh_inc data

growth_top4_sum_dbh<-ggplot(top4_dbh, aes(x=factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")), y=inc_annual,
                                              fill=factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")),
                                              color=factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))
                                              ))+ 
  geom_violin(position = "dodge")+ 
    geom_boxplot(position="dodge", col="black", fill="white", width=0.1)+
ylab("Annualised diameter increment (cm)")+
xlab("species")+theme_bw()+
  geom_text(data=top4_dbh %>% group_by(sp) %>% dplyr::summarise(n=n()),
   aes(x=factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")), y=2.2, 
  label=n))+
  geom_hline(yintercept = 0, lty=3)+
  ggtitle("tree growth in HKK from tape measure")+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  guides(fill="none",
         color="none")

png("growth-precip-thailand/doc/display/top4_increments_sum_dbh.png", width=4, height=4, units="in", res=300)
growth_top4_sum_dbh
dev.off()

```
```{r comparing dbh and dendro, echo=F}

#merge dbh and dendro data by Tag and Cno

inc_all<-merge(dendro_inc, dbh_inc, by=c("Tag", "Cno"), all=T)

#plot increments from dbh and dendro data

png("growth-precip-thailand/doc/display/dbh_dendro_compare.png", width=4, height=4, units="in", res=300)
ggplot(inc_all %>% filter(sp.x %in% c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")), aes(x=inc_annual.x, y=inc_annual.y, 
col=factor(sp.x, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))))+
  geom_point()+
  scale_color_viridis_d()+
  geom_abline(intercept=0, slope=1)+
  xlab("dendroband increment (cm)")+
  ylab("dbh increment (cm)")+
  guides(col=guide_legend(title="species"))+
  ggtitle("comparison of dbh and dendro increments")+
  theme_bw()
dev.off()
```

```{r top4 species growth contribution, echo=F}
